Learning \strips\ action models from {\em state-invaraints}

### 2019-02-04 doc/ijcai-19-deductivelearning/deductivelearning.tex
first commit ijcai19 paper on deductive learning

\section{Introduction}
\label{sec:introduction}
{\em Classical planing} is an interesting approach for learning \strips\ action models since it is flexible to different kinds of input knowledge (e.g., partially/fully observations of actions of plan executions as well as partially/fully observed intermediate states)~\cite{aineto2018learning}. This paper shows that this flexibility of the compilation approach goes one step further since it also allows learning from {\em state-invariants}, logic formulae that constrain the set of possible states of a given domain. The experimental results show that {\em state-invariants} boost the performance of the {\em classical planing} compilation for learning \strips\ action models.

@@

The formula $\phi_{I,A}^*$ represents the {\em strongest invariant} and exactly characterizes the set of all states reachable from $I$ with the actions in $A$. For instance Figure~\ref{fig:strongest-invariant} shows five clauses that define the {\em strongest invariant} for {\em blocksworld}. There are infinitely many strongest invariants, but they are all logically equivalent, and computing the strongest invariant is PSPACE-hard as hard as testing plan existence.

\begin{figure}[hbt!]
  \begin{footnotesize}
$\forall x_1,x_2\ ontable(x_1)\leftrightarrow\neg on(x_1,x_2)$.\\
$\forall x_1,x_2\ clear(x_1)\leftrightarrow\neg on(x_2,x_1)$.\\
$\forall x_1,x_2,x_3\ \neg on(x_1,x_2)\vee\neg on(x_1,x_3)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,x_2,x_3\ \neg on(x_2,x_1)\vee\neg on(x_3,x_1)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,\ldots,x_n\ \neg(on(x_1,x_2)\wedge on(x_2,x_3)\wedge\ldots\wedge on(x_{n-1},x_n)\wedge on(x_n,x_1)).$
\end{footnotesize}
 \caption{\small Example of the {\em strongest invariant} for the {\em blocksworld} domain.}
\label{fig:strongest-invariant}
\end{figure}

@@

\section{Deductive learning of planning action models}
\label{sec:evaluation}

The task of learning action models is defined by $\Lambda=\tup{\mathcal{M_0},\Phi,\mathcal{O}(\tau)}$:
\begin{itemize}
\item $\mathcal{M_0}$ is the {\bf initial domain model} (set of action models). This set is {\em empty}, when learning from scratch, or {\em partially specified}, when some fragments of the action models are known a priori.
\item $\Phi$, a set of {\em state-invariants} that constraint the set of possible states in the given domain.
\item $\mathcal{O}(\tau)$ is an observation of a trajectory $\tau(\pi,P)$ produced by the execution of an unknown plan $\pi$ that reaches the goals $G\in G[\cdot]$ starting from the initial state $I$ in $P$.
\end{itemize}

A {\em solution} to a learning task $\Lambda=\tup{\mathcal{M},\mathcal{O}(\tau)}$ is a domain model $\mathcal{M}'$ that is consistent with the information of $\mathcal{M}$ and with the observed plan trace $\tau$. This means that the action sequence (plan) that solves the planning problem $\tup {s_0,s_n}$ with $\mathcal{M}'$ along with the state trajectory induced by this plan is consistent with the input observatio $\mathcal{O}(\tau)$ and only travers states that satisfy the states invariants.

\subsection{Learning of planning action models with classical planning}
Because of the combinatorial nature of the search for a solution plan, the sooner unpromising nodes are pruned from the search the more efficient the computation of a solution plan. Constraints can be used to confine earlier the set of possible \strips\ action models and reduce then the learning hypothesis space. With regard to our compilation, {\em domain mutex} are useful to reduce the amount of applicable actions for programming a precondition or an effect for a given action schema. For example given the {\em domain mutex} $\phi=(\neg f_1\vee \neg f_2)$ such that $f_1\in F_v(\xi)$ and $f_2\in F_v(\xi)$, we can redefine the corresponding programming actions for {\bf removing} the {\em precondition} $f_1\in F_v(\xi)$ from the action schema $\xi\in\mathcal{M}$ as:

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programPre_{f_1,\xi}})=&\{\neg del_{f_1}(\xi),\neg add_{f_1}(\xi), mode_{prog}, pre_{f_1}(\xi), pre_{f_2}(\xi)\},\\
\cond(\mathsf{programPre_{f_1,\xi}})=&\{\emptyset\}\rhd\{\neg pre_{f_1}(\xi)\}.
\end{align*}
\end{small}

>>>>

\section{Introduction}
\label{sec:introduction}

The specification of planning action models is a complex task, even for planning experts, which frequently causes a knowledge acquisition bottleneck that limits the potential of AI planning~\cite{kambhampati:modellite:AAAI2007}. The {\em machine learning} of action models aims to relieve this knowledge acquisition bottleneck. Nowadays there is a wide range of different approaches for learning planning action models~\cite{arora:amodels:ker2018}. However, most of these approaches aim maximizing some measure of {\em statistical consistency} over large sets of input examples that typically are sequential plans.

{\em Classical planning} is an interesting approach for learning \strips\ action models since it is flexible to different kinds of input knowledge (e.g., partially/fully observations of actions of plan executions as well as partially/fully observed intermediate states) and guaranttees to output an action model that is {\em consistent} with the input knowledge~\cite{aineto2018learning}. In this paper we show that this flexibility of the compilation approach goes one step further since it also allows learning from {\em state-invariants}, logic formulae that constrain the set of possible states of a given domain. The experimental results show that {\em state-invariants} boost the performance of the {\em classical planing} compilation for learning \strips\ action models.

@@

The formula $\phi_{I,A}^*$ represents the {\em strongest invariant} and exactly characterizes the set of all states reachable from $I$ with the actions in $A$. For instance Figure~\ref{fig:strongest-invariant} shows five clauses that define the {\em strongest invariant} for the {\em blocksworld} planning domain~\cite{slaney2001blocks}. There are infinitely many strongest invariants, but they are all logically equivalent, and computing the strongest invariant is PSPACE-hard (as hard as testing plan existence~\cite{bylander:complexity:AIJ1994}).

\begin{figure}[hbt!]
  \begin{footnotesize}
$\forall x_1,x_2\ ontable(x_1)\leftrightarrow\neg on(x_1,x_2)$.\\
$\forall x_1,x_2\ clear(x_1)\leftrightarrow\neg on(x_2,x_1)$.\\
$\forall x_1,x_2,x_3\ \neg on(x_1,x_2)\vee\neg on(x_1,x_3)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,x_2,x_3\ \neg on(x_2,x_1)\vee\neg on(x_3,x_1)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,\ldots,x_n\ \neg(on(x_1,x_2)\wedge on(x_2,x_3)\wedge\ldots\wedge on(x_{n-1},x_n)\wedge on(x_n,x_1)).$
\end{footnotesize}
 \caption{\small {\em Strongest invariant} for the {\em blocksworld} domain.}
\label{fig:strongest-invariant}
\end{figure}

@@

\section{Learning action models from {\em state-invaraints}}
\label{sec:evaluation}

We define the task of learning action models as a tuple $\Lambda=\tup{M,\Phi,\mathcal{O}(\tau)}$, where:
\begin{itemize}
\item $M$ is the {\em space of possible action model}. This set is the {\em full} space of action models, when learning from scratch, or {\em partially specified}, when some fragments of the action models are known a priori. A set of action models can be defined {\em explicitly}, enumerating all the models that belong to the set or {\em implicitly}, enumerating all the constraints that must satisfy any model that belongs to the set. A {\em partially specified} \strips\ action model is then a formalism for the {\em implicit} representation of a set of \strips\ schemes~\cite{sreedharan2018handling}. 
\item $\Phi$, a set of {\em state-invariants} that constrain the set of possible states in the given domain.
\item $\mathcal{O}(\tau)$ is an observation of a trajectory $\tau(\pi,P)$ produced by the execution of an unknown plan $\pi$ that solves $P_\mathcal{O}$ (i.e. the classical planning problem $P_\mathcal{O}=\tup{F,A[\cdot],s_0^o,s_m^o}$ where $A[\cdot]$ is a set of actions s.t., the semantics of each action $a\in A[\cdot]$ is unknown since functions $\rho$ and/or $\theta$ of $a$ are undefined). 
\end{itemize}

A {\em solution} to a $\Lambda=\tup{M,\Phi,\mathcal{O}(\tau)}$ learning task is a model $\mathcal{M}'\in M$ that is {\em consistent} with the given input knowledge $\Phi$ and $\mathcal{O}(\tau)$. This means that let there exists a plan $\pi$ such that the observation $\mathcal{O}(\tau)$ is consistent with $\pi$ when the dynamics of $A[\cdot]$ is given by $\mathcal{M}'\in M$ and such that any state produced in the trajectory $\tau(\pi,P_\mathcal{O})$ satisfies the {\em state-invariants} $\Phi$.

Next we show that the set $M$ of possible action models can be encoded as a set of propositional variables and a set of constraints over those variables. Then, we show how to exploit this encoding to solve the $\Lambda=\tup{M,\Phi,\mathcal{O}(\tau)}$ learning task with an off-the-shelf classical planner.

\subsection{A propositional encoding for the space of STRIPS action models}
{\em A \strips\ action schema} $\xi$ is defined by four lists: A list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema. Let be $\Psi$ the set of {\em predicates} that shape the propositional state variables $F$, and a list of {\em parameters} $pars(\xi)$. The set of elements that can appear in $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the \strips\ action schema $\xi$ is given by FOL interpretations of $\Psi$ over the parameters $pars(\xi)$ and is denoted as ${\mathcal I}_{\Psi,\xi}$.

For instance in a four-operator {\em blocksworld}~\cite{slaney2001blocks}, the ${\mathcal I}_{\Psi,\xi}$ set contains only five elements for the {\small \tt pickup($v_1$)} schemata, ${\mathcal I}_{\Psi,pickup}$={\small\tt\{handempty, holding($v_1$), clear($v_1$), ontable($v_1$), on($v_1,v_1$)\}} while it contains eleven elements for the {\small \tt stack($v_1$,$v_2$)} schemata, ${\mathcal I}_{\Psi,stack}$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$), on($v_1,v_2$), on($v_2,v_1$), on($v_2,v_2$)\}}.

Despite any element of ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schema $\xi$, the actual space of possible \strips\ schemata is bounded by constraints of three kinds:
\begin{enumerate}
\item {\bf Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. {\em Typing constraints} are also of this kind~\cite{mcdermott1998pddl}. 
\item {\bf Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because, in this specific domain, a block cannot be on top of itself. {\it State invariants} are also constraints of this kind. 
\item {\bf Observation constraints}. An observation $\mathcal{O}(\tau)$ depicts {\em semantic knowledge} that constraints further the space of possible action schemata.   
\end{enumerate}

\begin{figure}
  \begin{tiny}  
  \begin{verbatim}
(:action stack
   :parameters (?v1 ?v2)
   :precondition (and (holding ?v1) (clear ?v2))
   :effect (and (not (holding ?v1)) (not (clear ?v2))
                (clear ?v1) (handempty) (on ?v1 ?v2)))

(pre_holding_v1_stack) (pre_clear_v2_stack)
(eff_holding_v1_stack) (eff_clear_v2_stack)
(eff_clear_v1_stack) (eff_handempty_stack) (eff_on_v1_v2_stack)
  \end{verbatim}           
  \end{tiny}  
 \caption{\small PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema.}
\label{fig:propositional}
\end{figure}

In this work we introduce a propositional encoding of the {\em preconditions}, {\em negative}, and {\em positive} effects of a \strips\ action schema $\xi$ using only fluents of two kinds {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} (where $e\in{\mathcal I}_{\Psi,\xi}$). This encoding exploits the syntactic constraints of \strips\, so it is more compact that the one previously proposed by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} for learning classical planning action models. In more detail, if {\tt\small pre\_e\_$\xi$} holds it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a {\em precondition} in $\xi$. If {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} holds it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a {\em negative effect} in $\xi$ while if $pre\_e\_\xi$ does not hold but {\tt\small eff\_e\_$\xi$} holds, it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a {\em positive effect} in $\xi$. Figure~\ref{fig:propositional} shows the PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema with {\tt\small pre\_e\_stack} and {\tt\small eff\_e\_stack} fluents ($e\in{\mathcal I}_{\Psi,stack}$).

\subsection{Learning \strips\ action models with classical planning}
Given a $\Lambda=\tup{M,\Phi,\mathcal{O}(\tau)}$ where $\Phi$ is a set of {\em domain mutex} $\phi\in\Phi$, we create a classical planning problem $P'=\tup{F',A',I,G}$ such that:
\begin{itemize}
\item $F'$ extends $F$ with a fluent $mode_{insert}$, to indicate whether action models are being programmed, and the fluents for the propositional encoding of the corresponding space of STRIPS action models. This is a set of fluents of the type $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of FOL interpretations of predicates $\Psi$ over the corresponding action parameters $pars(\xi)$.

\item $A'$ replaces the actions in $A$ with two types of actions.
\begin{enumerate}
\item Actions for {\em inserting} a {\em precondition}, {\em positive} effect or {\em negative} effect in $\xi$ following the syntactic constraints of \strips\ models. In the particular case that $M$ is a {\em partially specified model} then only the actions for inserting a possible {\em precondition} or {\em effect} are necessary.
\begin{itemize}
\item Actions which support the addition of a {\em precondition} $p\in \Psi_{\xi}$ to the action model $\xi$. A precondition $p$ is inserted in $\xi$ when neither $pre_p$, $eff_p$ exist in $\xi$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre_{p}(\xi), \neg eff_{p}(\xi), mode_{insert}\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre_{p}(\xi)\}.
\end{align*}
\end{small}

\item Actions which support the addition of a {\em negative} or {\em positive} effect $p\in \Psi_{\xi}$ to the action model $\xi$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertEff_{p,\xi}})=&\{\neg eff_{p}(\xi), mode_{insert}\},\\
\cond(\mathsf{insertEff_{p,\xi}})=&\{\emptyset\}\rhd\{eff_{p}(\xi)\}.
\end{align*}
\end{small}
\end{itemize}

\item Actions for {\em applying} an action model $\xi$ built by the {\em insert} actions and bounded to objects $\omega\subseteq\Omega^{|pars(\xi)|}$ (where $\Omega$ is the set of {\em objects} used to induce the fluents $F$ by assigning objects in $\Omega$ to the $\Psi$ predicates and $\Omega^k$ is the $k$-th Cartesian power of $\Omega$). The action parameters, $pars(\xi)$, are bound to the objects in $\omega$ that appear in the same position.
\end{enumerate}
\end{itemize}

\begin{small}
\begin{align*}
\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\implies p(\omega)\}_{\forall p\in\Psi_\xi},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi_\xi},\\
&\{\neg pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi\xi}\},\\
&\{\emptyset\}\rhd\{\neg mode_{insert}\}.
\end{align*}
\end{small}

{\em domain mutex} are useful to reduce the amount of applicable actions for programming a precondition or an effect for a given action schema. For example given the {\em domain mutex} $\phi=(\neg f_1\vee \neg f_2)$ such that $f_1\in F_v(\xi)$ and $f_2\in F_v(\xi)$, we can redefine the corresponding programming actions for {\bf removing} the {\em precondition} $f_1\in F_v(\xi)$ from the action schema $\xi\in\mathcal{M}$ as:
